---
title: "Quiz 1"
subtitle: "CSC 110"
author: "Your name here"
date: "Feb 06, 2025"
date-format: long
format:
  html:
    toc: true
    toc-depth: 2
    fig-width: 8
    fig-height: 4
  pdf:
    toc: true
    toc-depth: 1
    fig-width: 8
    fig-height: 5
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - heightrounded
---

# Instructions

Welcome to CSC 110 Quiz 1! In this lab, you will plot and analyze inequality and 
mobility data from Chetty using the ggplot2 package.

Quiz assignments can be taken **open-book** using the R4DS textbook and **open-note** 
using the notes you've taken and the materials we have covered in class. You may 
also use the built-in help features in RStudio. Otherwise, You should not rely 
on any outside resources. Usage of generative AI tools like ChatGPT, Claude, or 
DeepSeek is not allowed on this assignment.

-   You must work alone on this quiz. You may ask me a clarifying question by 
email or in office hours, but otherwise you are not to seek outside help on this 
assignment.
-   To earn full credit on coding questions, your code and its output must both 
be visible.
-   If a coding question produces a plot/figure:
    -   Include clear and descriptive labels (title, axis, etc.)
    -   Include the `label` and `fig-cap` options at the top of each code block
-   Written (non-code) answers to questions should be typed out in the answer 
boxes that are provided for you, and should be written in complete sentences.
Do not replace the heading in each answer box, but instead type your responses 
below each heading.
-   I recommend you save and render your Quarto doc as an HTML document 
regularly. This not only has the benefit of showing your output, but it also 
tests your code to make sure everything is working. Any errors that the 
rendering engine encounters will be displayed in the RStudio console.
-   **Remember:** all work is done under the Honor Code, and you must pledge 
your work at the end of the assignment.
-   When finished, render your Quarto doc as a PDF, review it carefully for 
completeness and cleanliness, and submit it in Gradescope. Your Qmd file will 
stay stored in JupyterHub if we need to refer back to it.

# Setup

Run the block below to load the necessary packages for this quiz.

```{r setup}
#| label: load-packages
#| include: true
#| message: false

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

\newpage

# Questions

## Question 1.1

Write an R statement that creates a data frame named "utip" from the Excel file 
contained in utip_plus.xlsx. Then use one of the techniques we've learned to 
preview the data so you can answer the question below.

```{r}
#| label: "Question 1.1"


```

## Question 1.2

How many rows are in this data frame? What does each row represent?

::: {.callout-note title="Answer to 1.2" appearance="minimal"}
(replace this with your answer)
:::

\newpage

## Question 2.1

Use your knowledge of ggplot to produce a scatterplot that shows the relationship 
between tract-to-tract income inequality (`tract_gini`) and 
household-to-household income inequality (`hh_gini`), with each point marked by 
a filled blue square.

```{r}
#| label: "Question 2.1"
#| fig-cap: "Plot 2.1 - tract-to-tract inequality vs. household-to-household inequality"

```

## Question 2.2

What does each point represent in the plot?

::: {.callout-note title="Answer to 2.2" appearance="minimal"}
(replace this with your answer)
:::

\newpage

## Question 3.1

Produce a scatterplot that shows the relationship between household-to-household 
income inequality (`hh_gini`) and absolute mobility (`avg_mob`).

```{r}
#| label: "Question 3.1"
#| fig-cap: "Plot 3.1 - household-to-household inequality vs. absolute mobility"

```

## Question 3.2

Do you see any trends? If so, describe them. Are these relationships what you 
might have expected based on the Chetty study?

::: {.callout-note title="Answer to 3.2" appearance="minimal"}
(replace this with your answer)
:::

\newpage

## Question 4.1

Explore the potential effects of Simpson's Paradox in the relationship shown 
in Question 3.1 by adding aesthetics and/or faceting on the variables `inctax`
and `region.`

```{r}
#| label: "Question 4.1"
#| fig-cap: "Plot 4.1 - Exploring Simpson's Paradox"

```

## Question 4.2

Describe your findings, and be specific. Do you see evidence of Simpson's Paradox here?

::: {.callout-note title="Answer to 4.2" appearance="minimal"}
(replace this with your answer)
:::

\newpage

## Question 5.1

Produce a bar chart that visualizes the number of states in each region and in 
each income tax category.

```{r}
#| label: "Question 5.1"
#| fig-cap: "Plot 5.1 - States by region and income tax category"

```

## Question 5.2

Explain your choice of x and fill aesthetics, and your choice of position.

::: {.callout-note title="Answer to 5.2" appearance="minimal"}
(replace this with your answer)
:::

# Honor Code Pledge

On my honor I have neither given nor received unauthorized information regarding this work, I have followed and will continue to observe all regulations regarding it, and I am unaware of any violation of the Honor Code by others.

(Type your full name below to pledge this assignment.)

::: {.callout-note title="Honor Code Pledge" appearance="minimal"}
(replace this with your full name to pledge this assignment)
:::

# Submission checklist

Before submitting your assignment in Gradescope, review this checklist carefully 
to ensure you have done everything appropriately.

-   Correct author and date
-   All code is clearly visible
-   All output is clearly visible (does not apply to the Setup section), 
including labels and legends
-   Written responses are typed into the appropriate answer boxes
-   All empty or redundant plots are removed
-   Honor Code pledge is signed

Point penalties will be assessed if your assignment does not meet these guidelines.

\newpage

# Grading

### Setup question

### Data viz questions

-   1 point = Code and output are both clearly and completely visible
-   1 point = Correct geom is used
-   1 point = Correct aesthetics are applied
-   1 point = Descriptive labels are applied

### Short answer questions

-   1 point = Answers the question thoughtfully
-   1 point = Provides clear expository writing
